\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-28 papers/jfp/3-abstractions.tex
Polishing S1-S3
Polishing S3-S4

% \vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
% \vspace{-0.5mm}

@@

\emph{Build system.} A \emph{build system} takes a task description, a
\emph{target} key, and a store, and returns a new store in which the target key
and all its dependencies have an up to date value.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task  c k v = Task
    { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}

@@

Nevertheless, \hs{busy} can easily handle the example \hs{sprsh1}
from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
initialise the store with \cmd{A1} set to 10 and all other cells set to 20.

@@

The previous example shows why the \hs{Task} abstraction is polymorphic in
\hs{f}, recall the definition:

\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
\end{minted}

@@

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
systems that can build \cmd{T} -- and we will do so in
section~\S\ref{sec-implementations}. As we shall see, each build system will use
a different \hs{f}, so the task description must not fix \hs{f}.

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}. So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  linear chain, as e.g. in Docker containers~\cite{docker}. It is
  interesting to note that, when run on such a task description, the \hs{busy}
  build system will build each key at most once, thus partially fulfilling the
  minimality requirement~\ref{def-minimal}. Alas, it still has no mechanism to
  decide which input keys changed since the previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\noindent
Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, but \Excel
(which uses dynamic dependencies) is perfectly happy. We can express this
spreadsheet using our task abstraction as:

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Tasks Monad String Integer
sprsh2 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 "B2" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _ = Nothing
\end{minted}
\vspace{0.5mm}

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task $
    \k -> Identity $ getValue k store
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{\$}~\hs{getValue}~\hs{k}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result has type \hs{Identity}~\hs{v}, which we unwrap with \hs{runIdentity}.

% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type
% \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
% which both forces the monad to \hs{Identity} and strips it away.

% \vspace{-1mm}
\textbf{Definition}~(Correctness). Suppose \hs{build} is a build system,
\hs{tasks} is a build task description, \hs{key} is a target key, \hs{store} is
an initial store, and \hs{result} is the store produced by running the build
system with parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the
precise language of our abstractions:

@@

    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    corresponding \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{tasks}, \hs{key} and \hs{store}.

\label{def-correct}
% \vspace{2mm}

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is extremely cool. We will see a practical
use for \hs{dependencies} when implementing applicative build systems,
see~\S\ref{sec-implementation-make}.

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
on a store with concrete values, which will determine the discovered
dependencies. Accordingly, we introduce the function \hs{track}: a combination
of \hs{compute} and \hs{dependencies} that computes both the resulting value and
the list of its dependencies (key/value pairs) in an arbitrary monadic
context~\hs{m}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
track task fetch = runWriterT $ run task trackingFetch
  where
    trackingFetch :: k -> WriterT [(k, v)] m v
    trackingFetch k = do v <- lift (fetch k); tell [(k, v)]; return v
\end{minted}
\vspace{1mm}

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us to record additional
information -- a list of key/value pairs \hs{[(@@k,}~\hs{v)]} -- when computing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,[("C1",1),("B2",10)])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,[("C1",2),("A2",20)])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
which in this case is obtained by reading from the standard input.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependency :: Task Functor k v -> k
dependency task = getConst $ run task Const
\end{minted}
\vspace{1mm}

>>>>

\vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
\vspace{-0.5mm}

@@

\emph{Build system.} A \emph{build system} takes a task description, a
\emph{target} key, and a store, and returns a new store in which the target key
and all its dependencies have up to date values.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task  c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}

@@

Nevertheless, \hs{busy} can easily handle the example task description
\hs{sprsh1} from the previous subsection~\S\ref{sec-task}. In the GHCi session
below we initialise the store with \cmd{A1} set to~10 and all other cells set
to~20.

@@

The previous example illustrates why the \hs{Task} abstraction is polymorphic
in~\hs{f}; recall its definition from~\S\ref{sec-task}:

\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task c k v = Task@\,@{@\,\blk{run}\,@::@\,@@\std{forall}@ f. c f => (k -> f v) -> f v@\,@}
\end{minted}

@@

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
systems that can build \cmd{T} -- and we will do so in
section~\S\ref{sec-implementations}. As we shall see, each build system will use
a different \hs{f}, so the task description must not fix~\hs{f}.

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}. So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\noindent
Note that \cmd{B1} and \cmd{B2} statically form a dependency cycle, so \Make
would not be able to order the tasks topologically, but \Excel, which uses
dynamic dependencies, is perfectly happy. We can express this spreadsheet using
our task abstraction as follows:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Tasks Monad String Integer
sprsh2 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 "B2" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity (run task fetch)
  where
    fetch :: k -> Identity v
    fetch k = Identity (getValue k store)
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
we unwrap with \hs{runIdentity}.

\vspace{2mm}
\textbf{Definition}~(Correctness). Suppose \hs{build} is a build system,
\hs{tasks} is a build task description, \hs{key} is a target key, \hs{store} is
an initial store, and \hs{result} is the store produced by running the build
system with parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the
precise language of our abstractions:

@@

    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    corresponding \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
\hs{tasks}, \hs{key} and \hs{store}.

\label{def-correct}
\vspace{2mm}

@@

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~Capriotti~\&~Kaposi~\shortcite{free-applicatives}, but it is extremely cool.
We will see a practical use for \hs{dependencies} when implementing applicative
build systems, see~\S\ref{sec-implementation-make}.

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
on a store with concrete values, which will determine the discovered
dependencies. Accordingly, we introduce the function \hs{track} -- a combination
of \hs{compute} and \hs{dependencies} that computes both the resulting value and
the list of its dependencies (key/value pairs) in an arbitrary monadic
context~\hs{m}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer

track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
track task fetch = runWriterT $ run task trackingFetch
  where
    trackingFetch :: k -> WriterT [(k, v)] m v
    trackingFetch k = do v <- lift@\,@(fetch k); tell@\,@[(k, v)]; return@\,\blk{v}@
\end{minted}
\vspace{1mm}

\noindent
This function uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad} for recording additional
information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,[("C1",1),("B2",10)])
\end{minted}
\vspace{-2mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,[("C1",2),("A2",20)])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of the cell \cmd{B1} from \hs{sprsh2} (see the
spreadsheet in~\S\ref{sec-task-monad}) are determined by the value of \cmd{C1},
which in this case is obtained by reading from the standard input using
\hs{fetchIO}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependency :: Task Functor k v -> k
dependency task = getConst (run task Const)
\end{minted}
\vspace{1mm}

### 2019-04-28 papers/jfp/4-schedulers.tex
Start work on rebuilders
Polishing S3-S4

\section{Build System Schedulers}\label{sec-scheduler}

The focus of this paper is on a variety of implementations of
\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
a \emph{client-supplied} implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That
is, we are going to take \hs{Tasks} as given from now on, and explore variants of
\hs{Build}: first abstractly (in this section) and then concretely
in~\S\ref{sec-implementations}.

@@

We have emboldened two different aspects above: the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). In this section we discuss schedulers,
leaving rebuilders for \S\ref{sec-rebuilder}.

@@

The topological scheduler pre-computes a linear order of tasks, which when followed ensures
dependencies are satisifed, then executes the required tasks in that order. Computing such a linear
order is straightforward -- given a task description and output \hs{key} first find the (acyclic)
graph of the \hs{key}'s dependencies, then compute a topological sort. Taking the example from
Figure~\ref{fig-make}, we might compute the order:

\begin{enumerate}
\item \hs{util.o}
\item \hs{main.o}
\item \hs{main.exe}
\end{enumerate}

Given the dependencies, we could have equally chosen to build \hs{main.o} first, but \hs{main.exe} \emph{must}
come last.

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key \hs{dep}
abort the task and build \hs{dep} instead. Returning to the example from
Figure~\ref{fig-make}, we might build in the order:

\begin{enumerate}
\item \hs{main.exe} (abort because it requires \hs{util.o})
\item \hs{util.o}
\item \hs{main.o}
\item \hs{main.exe}
\end{enumerate}

Here we start with \hs{main.exe} (chosen randomly), but discover it requires \hs{util.o}, so
instead start building \hs{util.o}. Next we choose to build \hs{main.o} (again, randomly),
before finally returning to \hs{main.exe} which now has all its dependencies available and
completes without error.

The approach works, but has a number of disadvantages. Firstly, it requires a technical mechanism
to abort a task -- easy enough in our theoretical setting of \hs{Task} but leads to engineering concerns
in the real world. Secondly, it is not minimal in the sense that a task may
start, do some meaningful work, and then abort, repeating that same work when restarted.

As a refinement, to reduce the number of aborts (often to zero) \Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information from which it can computed a linear order. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is out of date.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task. Using
Figure~\ref{fig-make}, we would build:

\begin{enumerate}
\item \hs{main.exe} (suspended)
\begin{itemize}
\item \hs{util.o}
\end{itemize}
\item \hs{main.exe} (resumsed then suspended)
\begin{itemize}
\item \hs{main.o}
\end{itemize}
\item \hs{main.exe} (completed)
\end{enumerate}

As we see, the \hs{main.exe} is started first as it is the required target. It soon discovers
a dependency on \hs{util.o} and suspends to build \hs{util.o}, resumes a build before building \hs{main.o}
then continues its work to finish \hs{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal build system
which supports dynamic dependencies.

The difficulties are all engineering in nature. To implement suspension there are two
standard approaches:

@@

While a suspending scheduler is theoretically optimal, it is only better in practice
than a restarting scheduler if the cost of duplicate work avoided outweighs the cost of suspending tasks.

>>>>

\section{Schedulers}\label{sec-scheduler}

The focus of this paper is on a variety of implementations of
\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given a \emph{user-supplied}
implementation of \hs{Tasks}~\hs{c}~\hs{k}~\hs{v}. That is, we are going to take
\hs{Tasks} as given from now on, and explore variants of \hs{Build}: first
abstractly (in this section and in~\S\ref{sec-rebuilder}) and then concretely
in~\S\ref{sec-implementations}.

@@

We have emboldened two different aspects above: the part of the build system
responsible for scheduling tasks in the dependency order (a ``scheduler'') can
be cleanly separated from the part responsible for deciding whether a key needs
to be rebuilt (a ``rebuilder''). In this section we discuss schedulers, leaving
rebuilders for \S\ref{sec-rebuilder}.

@@

The topological scheduler pre-computes a linear order of tasks, which when
followed ensures dependencies are satisfied, then executes the required tasks in
that order. Computing such a linear order is straightforward -- given a task
description and output \hs{key} first find the (acyclic) graph of the \hs{key}'s
dependencies, then compute a topological order. Taking the \Make example from
Fig.~\ref{fig-make}, we might compute the order:

\begin{enumerate}
\item \cmd{util.o}
\item \cmd{main.o}
\item \cmd{main.exe}
\end{enumerate}

\noindent
Given the dependencies, we could have equally chosen to build \cmd{main.o}
first, but \cmd{main.exe} \emph{must} come last.

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key
\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
from Fig.~\ref{fig-make}, we might build in the order:

\begin{enumerate}
\item \cmd{main.exe} (abort because it requires \cmd{util.o})
\item \cmd{util.o}
\item \cmd{main.o}
\item \cmd{main.exe} (restart from scratch, completing successfully this time)
\end{enumerate}

\noindent
We start with \cmd{main.exe} (chosen randomly), but discover it requires
\cmd{util.o}, so instead start building \cmd{util.o}. Next we choose to build
\cmd{main.o} (again, randomly), before finally returning to \cmd{main.exe},
which now has all its dependencies available and completes without error.

The approach works, but has a number of disadvantages. Firstly, it requires a
technical mechanism to abort a task -- easy enough in our theoretical setting of
\hs{Task} (see the implementation in~\S\ref{sec-implementation-excel}) but leads
to engineering concerns in the real world. Secondly, it is not minimal in the
sense that a task may start, do some meaningful work, and then abort, repeating
that same work when restarted.

As a refinement, to reduce the number of aborts (often to zero) \Excel records
the discovered task order in its \emph{calc chain}, and uses it as the starting
point for the next build~(\S\ref{sec-background-excel}). \Bazel's restarting
scheduler does not store the discovered order between build runs; instead, it
stores the most recent task dependency information from which it can compute a
linear order. Since this information may become outdated, \Bazel may also need
to abort a task if a newly discovered dependency is out of date.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task. Using the example
from Fig.~\ref{fig-make}, we would build:

\begin{itemize}
\item \cmd{main.exe} (suspended)\\
  $\hookrightarrow$ \cmd{util.o}
\item \cmd{main.exe} (resumed then suspended again)\\
  $\hookrightarrow$   \cmd{main.o}
\item \cmd{main.exe} (completed)
\end{itemize}

\noindent
As we see, the \cmd{main.exe} is started first as it is the required target. It
soon discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
resumes a build before building \cmd{main.o} then continues its work to finish
\cmd{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal
build system which supports dynamic dependencies. The difficulties are all
engineering in nature. To implement suspension there are two standard
approaches:

@@

\noindent
While a suspending scheduler is theoretically optimal, it is only better in
practice than a restarting scheduler if the cost of avoided duplicate work
outweighs the cost of suspending tasks. Note furthermore that the cost of
duplicate work may often be just a fraction of the overall build cost.

\q{Andrey}{It's unclear how our \hs{busy} build system fits into the two above
approaches. It probably is a third approach (i.e. using stack), but it's
unsuitable for real build systems that call external processes. Shall we mention
this?}

\q{Andrey}{A follow-up to the above: our \Shake models use neither of the two
above approaches, which seems a bit unsatisfactory. Can we at least hint how
our models could be adapted to use the continuation-passing style? Can we write
\hs{busyCPS}?}
